---
title: "Independence of Age Group and Shelf Location"
author: "Jorge Mendez"
date: "2025-11-20"
output: html_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```


## Purpose of the Lab


In this lab, we study whether the target consumer of a cereal (children vs adults) is related to the shelf position where the cereal is placed (bottom, middle, top).

We will:

  - Summarize the relationship between target consumer and shelf   location with a table.

  - Use a chi-square test of independence to see if the variables are related.

  - Look at the expected counts to understand any warnings from R.

  - Perform a permutation test for independence and compare the result with the chi-square test.

```{r}
library(resampledata3)
data(Cereals)

```

## (a) Contingency Table: Target Consumer × Shelf Location

First, we create a table that summarizes how many cereals for children and adults are placed on each shelf (bottom, middle, top).

```{r}
tbl <- table(Cereals$Age, Cereals$Shelf)
tbl


```
To see the totals for rows and columns:
```{r}
addmargins(tbl)

```

### Interpretation

This table shows how cereals are distributed across shelf positions for the different age target groups in the dataset.
From the counts (and the row/column totals), we can see whether:

   - Certain age groups (younger vs older consumers) have cereals placed more often on a particular shelf, and

   - Whether shelf placement is evenly distributed across age groups or shows a pattern.
   
This summary helps us evaluate whether Age and Shelf location are independent.

## (b) Chi-Square Test of Independence

Now that we have the contingency table, we use R’s chisq.test() function to test whether Age and Shelf location are independent.

The hypotheses are:

We test:

\[
H_0: \text{Age group and Shelf location are independent}
\]

\[
H_A: \text{Age group and Shelf location are not independent}
\]


The chi-square statistic is computed as

\[
\chi^2 = \sum_{\text{all cells}} \frac{(O - E)^2}{E},
\]

where \(O\) is the observed count and \(E\) is the expected count under independence.


```{r}
chisq_test <- chisq.test(tbl)
chisq_test

```

### Interpretation

R outputs the chi-square test statistic, degrees of freedom, and the p-value.

  - A large chi-square statistic and

  - A small p-value

provide evidence that Age group and Shelf location are related.

This test tells us whether the differences in counts across shelves are more than what we would expect by random chance.


## (c) Expected Counts and Warning Message

R gives the warning: “Chi-squared approximation may be incorrect.”

This usually happens when some of the expected counts in the table are too small.
We can look at the expected values directly:


The expected count for cell \((i,j)\) is

\[
E_{ij} = \frac{(\text{row total})_i \times (\text{column total})_j}{n}.
\]

We can look at the expected values directly:

```{r}
chisq_test$expected
```


### Interpretation

From the expected table, we see that one of the cells has an expected count below 5:

  - Adult & Bottom shelf → 3.56, which is less than 5

The chi-square test assumes that expected counts should generally be at least 5, or that at least 80% of the cells are ≥ 5.

Because this condition is violated, the chi-square p-value may not be fully reliable.
This is why we will perform a permutation test in part (d), which does not rely on this assumption.


## (d) Permutation Test of Independence

Because the chi-square test gave a warning about small expected counts, we use a permutation test as a non-parametric alternative.
This test does NOT rely on large expected frequencies, so it gives a more reliable answer.

The steps:

   1.- Compute the observed chi-square statistic

   2.- Shuffle the Age labels many times

   3.- Recompute the chi-square statistic for each shuffled dataset

   4.- Compare the observed statistic to the permutation distribution to get a p-value

The permutation p-value is estimated as

\[
\hat{p} = \frac{\#\{\text{permuted statistics } \ge \text{ observed statistic}\}}{\text{number of permutations}}.
\]


```{r}
set.seed(123)
n_perm <- 10000
perm_stats <- numeric(n_perm)

obs_stat <- chisq_test$statistic

for (i in 1:n_perm) {
shuffled_age <- sample(Cereals$Age)
tbl_perm <- table(shuffled_age, Cereals$Shelf)
perm_stats[i] <- suppressWarnings(chisq.test(tbl_perm)$statistic)
}

p_value_perm <- mean(perm_stats >= obs_stat)
p_value_perm

```


### Interpretation

The permutation p-value was 0, meaning that out of 10,000 shuffled datasets, none produced a chi-square statistic as large as the observed one. This implies

\[
p < \frac{1}{10{,}000} = 0.0001,
\]

which provides very strong evidence that Age and Shelf location are not independent.


Therefore, we conclude that Age group and Shelf location are not independent.
Cereals targeted at different age groups tend to be placed on different shelves, and this association is far stronger than what would be expected by chance alone.


## Conclusion

In this lab, we examined whether Age group and Shelf location are independent in the cereal dataset.
We began by creating a contingency table, which showed clear differences in shelf placement across age groups.
The chi-square test of independence produced a very small p-value, suggesting a strong association, but also generated a warning because one expected cell count was below 5.

To address this limitation, we conducted a permutation test, which does not rely on the large-sample assumptions required by the chi-square test.
After running 10,000 permutations, the permutation p-value was 0, meaning that none of the shuffled datasets produced a chi-square statistic as large as the observed one, providing extremely strong evidence against independence.


Overall, we conclude that Age group and Shelf location are not independent.
Cereals targeted to different age groups tend to be placed on different shelves, showing a strong and statistically significant association.

